## Important! Label and structure folders and xls. files as instructed below:
Layer 0 folder (root folder): E.g. R_Template_challenging_data
Layer 1 folders: animalID
Layer 2 xls. files: 

 i) Testis files:
  ia) Testis animalID (all round cells) (gel number)
  ib) Testis animalID (PS) (gel number)
  
 ii) Liver animalID (gel number)

# Follow the same folder and label structure as the provided training set.


```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "C:/Users/xima/Desktop/R_Template_challenging_data") # Change working directory to the layer 0 folder location (play button #1) or manually set working directory (follow provided word file for instruction and do not press play button #1).
```

```{r, message=FALSE, warning=FALSE, results='hide'}
## Run all libraries (Play button #2)

library(janitor)
library(stringr)
library(readxl)
library(stringi)
library(mclust)
library(tidyverse)
library(knitr)
library(dplyr)
library(ggplot2)
library(gridExtra)


```

```{r, warning=FALSE, message=FALSE, include=FALSE}

## Import raw data .xls-files for all cell types of each animal in each experiment (play button 3).

rm(list = ls())

testitem <- "Chemical_x" # Match the label of layer 1 folder. 

searchstring <- paste0("\\./",testitem)

organ <- "Testis" 

getwd()

dirlist <- tibble(list.dirs()) %>% 
  clean_names() %>% 
 dplyr::filter(grepl(searchstring,list_dirs))

dirlist <- dirlist[-1,]

i <- 1    
j <- i

for(i in seq_along(c(1:nrow(dirlist)))){
    dirnm <- dirlist$list_dirs[[i]]
  filelist <- tibble(list.files(dirnm)) %>% 
    rename(file_name = 1) %>%
    mutate(dirnme = dirnm) %>%
    mutate(spor = paste0(dirnme,"/", file_name)) %>%
    dplyr::filter(grepl(".xls",file_name))
  for(j in seq_along(1:nrow(filelist))){
    fn <- filelist$spor[j]
    flm <- filelist$file_name[j]
    dt <- read_excel(paste0(fn)) %>% 
      dplyr::select(1:8) %>% 
      clean_names()
    print(fn)
    dtf <- dt %>% 
      mutate(film = flm) %>%
      mutate(fileid = flm) %>% 
      mutate(PS=ifelse(grepl('PS',film), 'PS', "")) %>% 
      separate(film, c("tissue", "rat","slutten"), " ") %>%
      dplyr::select(!c(slutten)) %>%
      mutate(tstitem = testitem) %>% 
      mutate(sti = dirnm)
    
    ifelse(exists("All_comets"),
           All_comets <- rbind(All_comets,dtf),
           All_comets <- dtf)
  }
}

All_cometsRAW <- All_comets

All_comets <- unique(All_cometsRAW) %>%
  mutate(radnr = row_number()) %>%
  mutate(organ = tissue) %>%
  mutate(tissue=ifelse(grepl('PS',fileid), 'Testis_PS', as.character(tissue))) %>%
  mutate(tissue=ifelse(grepl('s\\(a',tissue), 'Testis', as.character(tissue))) %>%
  separate(sti, c("dot", "cc","AnimalID"), "/") %>%
  dplyr::select(!c(dot,cc))

```

```{r}
## Create distribution plots for each animal (play button #4).

testis_comets_testis <- All_comets %>% 
  dplyr::filter(tissue == "Testis")

testis_samplenames_testis <- unique(testis_comets_testis$AnimalID)

unique_animal_ids_testis <- unique(testis_comets_testis$AnimalID)

for (animal_id in unique_animal_ids_testis) {
  subset_data_testis <- testis_comets_testis[testis_comets_testis$AnimalID == animal_id, ]
  
  num_variables_testis <- nrow(subset_data_testis)
  
  hist(subset_data_testis$total_intensity, breaks = 40, freq = F,
       main = paste("Testis (all round cells), raw data", animal_id, ":", num_variables_testis, "comets"), xlab = "DNA content (Tot_Int)")
  
  title(sub = "Inspect the raw testis (all round comets) distrubution plot to \nremove non-relevant datapoints (debris and >4C population).", cex.sub = 0.5, line = 2.15, col.sub = "blue")
  
  helping_lines <- c(50000, 75000, 100000, 125000, 150000, 175000, 200000, 225000, 250000, 275000, 300000, 325000, 350000, 375000, 400000, 425000, 450000, 475000, 500000, 525000, 550000, 575000, 600000, 625000, 650000, 675000, 700000) # Options: Add or remove Tot_Int helping lines.
  
  for (i in seq_along(helping_lines)) {
    abline(v = helping_lines[i], col = "blue", lwd = 1)
  }
}
```


```{r}

## Export the raw data (File name: ”All_comets_raw”; median %TI, # comets) for all cell types of each animal to a txt-file (play button #5).
All_comets_raw <- All_comets %>%
  group_by(tissue, rat) %>%
  summarise(
    median_tail_intensity = median(tail_intensity, na.rm = TRUE),
    count_rows = n()
  )

output_All_comets_raw <- paste0("All_comets_raw_", testitem, ".txt")

write.table(All_comets_raw, file = output_All_comets_raw, sep = "\t", row.names = FALSE, quote = FALSE)

```


```{r}
## Inspect the order of your folder structure for each animal. Strictly follow this order for the subsequent operations (play button #6).


unique_animal_ids <- unique(testis_comets_testis$AnimalID)

print(unique_animal_ids)

```

```{r}
## Inspect the testis (all round comets) distribution plots generated in e) to remove the non-relevant datapoints (>4C population) based on Tot_Int for each animal.

## Insert (or replace) the >4C Tot_Int cut-off values of your animals in the correct order to remove the non-relevant datapoints >4C-population (play button #7).

conditions_lookup <- data.frame(
  AnimalID = unique(All_comets$AnimalID),
  remove4C_condition = c(240000, 260000), # Examples are used here. Replace with your >4C cut-off values.
  removeDebris_condition = c(50000, 50000) # Examples are used here. Replace with your debris cut-offs values.
)


all_comets_no_debris <- data.frame()

testis_animal_ids <- unique(All_comets$AnimalID[All_comets$tissue == "Testis"])
for (id in testis_animal_ids) {
  dt1 <- All_comets %>%
    filter(AnimalID == id & tissue == "Testis") %>%
    filter(total_intensity < conditions_lookup$remove4C_condition[conditions_lookup$AnimalID == id]) %>%
    filter(total_intensity > conditions_lookup$removeDebris_condition[conditions_lookup$AnimalID == id])
  all_comets_no_debris <- bind_rows(all_comets_no_debris, dt1)
}
```

```{r}

## Generate distribution plots all round comets, liver and PS (play button #8).

liver_comets <- All_comets %>% dplyr::filter(tissue == "Liver")
haploid <- all_comets_no_debris %>% dplyr::filter(tissue == "Testis")
testis_comets_PS <- All_comets %>% dplyr::filter(PS == "PS")

unique_animal_ids <- intersect(intersect(unique(liver_comets$AnimalID), unique(haploid$AnimalID)), unique(testis_comets_PS$AnimalID))

plot_histograms <- function(animal_id) {
  
  subset_data_haploid <- haploid %>% filter(AnimalID == animal_id)
  subset_data_liver <- liver_comets %>% filter(AnimalID == animal_id)
  subset_data_PS <- testis_comets_PS %>% filter(AnimalID == animal_id)
  
  num_variables_haploid <- nrow(subset_data_haploid)
  num_variables_liver <- nrow(subset_data_liver)
  num_variables_PS <- nrow(subset_data_PS)
  
  x_min <- min(c(subset_data_haploid$total_intensity, subset_data_liver$total_intensity, subset_data_PS$total_intensity), na.rm = TRUE)
  x_max <- max(c(subset_data_haploid$total_intensity, subset_data_liver$total_intensity, subset_data_PS$total_intensity), na.rm = TRUE)
  

  x_breaks <- seq(from = floor(x_min / 40000) * 40000, to = ceiling(x_max / 40000) * 40000, by = 40000)
  
  # Testis all round cells excluding debris/4N 
  p1 <- ggplot(subset_data_haploid, aes(x = total_intensity)) +
    geom_histogram(bins = 60, fill = "lightblue", color = "black") +
    labs(title = paste("All round cells,\nexcluding debris/4N\nAnimal ID:", animal_id, "\nComets:", num_variables_haploid),
         x = "DNA content (Tot_Int)", y = "Frequency") +
    scale_x_continuous(breaks = x_breaks) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
  
  # Liver
  p2 <- ggplot(subset_data_liver, aes(x = total_intensity)) +
    geom_histogram(bins = 60, fill = "darkred", color = "black") +
    labs(title = paste("Liver\nAnimal ID:", animal_id, "\nComets:", num_variables_liver),
         x = "DNA content (Tot_Int)", y = "Frequency") +
    scale_x_continuous(breaks = x_breaks) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
  
  # PS
  p3 <- ggplot(subset_data_PS, aes(x = total_intensity)) +
    geom_histogram(bins = 60, fill = "darkgreen", color = "black") +
    labs(title = paste("PS\nAnimal ID:", animal_id, "\nComets:", num_variables_PS),
         x = "DNA content (Tot_Int)", y = "Frequency") +
    scale_x_continuous(breaks = x_breaks) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
  
  grid.arrange(p1, p3, p2, ncol = 3)
}

for (animal_id in unique_animal_ids) {
  plot_histograms(animal_id)
}
```


```{r}

## Inspect the order of your folder structure for each animal. Follow this order for the subsequent operations (play button #9).
print(unique_animal_ids)
```

```{r}

## Manually insert the Tot_Int threshold values for the 1C population for each animal in the correct order. Separate the 1C population from the >1C populations (play button #10).



thresholds_visual <- data.frame(AnimalID = unique(testis_comets_testis$AnimalID),
                                Threshold = c(100000, 110000)) #Examples are used here. Replace with your Tot_Int threshold value for each animal. 

testis_comets_testis_threshold <- left_join(testis_comets_testis, thresholds_visual, by = "AnimalID")

haploid_visual <- testis_comets_testis_threshold %>%
  filter(total_intensity < Threshold)

all_round_rest_visual <- testis_comets_testis_threshold %>%
  filter(total_intensity > Threshold)

```

```{r}
## Export the 1C population data (File name: “1C_manual”; median %TI, # comets) of each animal to a txt file (play button #11).

animal_ids_all_round_visual <- unique(haploid_visual$AnimalID)

haploid_visual_inspection <- data.frame(AnimalID = character(), CometCount = numeric(), MedianTailIntensity = numeric(), stringsAsFactors = FALSE)

for (animal_id in animal_ids_all_round_visual) {
  animal_data <- haploid_visual[haploid_visual$AnimalID == animal_id, ]
  
  median_tail_intensity <- median(animal_data$tail_intensity)
  
  animal_count_all_round_visual <- sum(haploid_visual$AnimalID == animal_id)
  
  haploid_visual_inspection <- rbind(haploid_visual_inspection, data.frame(AnimalID = animal_id, CometCount = animal_count_all_round_visual, MedianTailIntensity = median_tail_intensity))
}

output_file_haploid_visual <- paste0("1C_manual_", testitem, ".txt")

write.table(haploid_visual_inspection, file = output_file_haploid_visual, sep = "\t", row.names = FALSE, quote = FALSE)

```

```{r}

## Export the >1C population data (File name: “More_than_1C_manual”; median %TI, # comets) of each animal to a txt file (play button #12). 
animal_ids_all_round_rest_visual <- unique(all_round_rest_visual$AnimalID)

all_round_rest_visual_inspection <- data.frame(AnimalID = character(), CometCount = numeric(), MedianTailIntensity = numeric(), stringsAsFactors = FALSE)

for (animal_id in animal_ids_all_round_rest_visual) {
  animal_data <- all_round_rest_visual[all_round_rest_visual$AnimalID == animal_id, ]
  
  median_tail_intensity <- median(animal_data$tail_intensity)
  
  animal_count_all_round_rest_visual <- sum(all_round_rest_visual$AnimalID == animal_id)
  
  all_round_rest_visual_inspection <- rbind(all_round_rest_visual_inspection, data.frame(AnimalID = animal_id, CometCount = animal_count_all_round_rest_visual, MedianTailIntensity = median_tail_intensity))
}

output_file_all_round_rest_visual_inspection <- paste0("More_than_1C_manual_", testitem, ".txt")

write.table(all_round_rest_visual_inspection, file = output_file_all_round_rest_visual_inspection, sep = "\t", row.names = FALSE, quote = FALSE)
```

```{r}

## Create distribution plots for each animal using Gaussian mixture model with three forced distributions (play button #13). The distribution plots include median %TI + 1 and 2 standard deviations of the first population.

unique_animal_ids <- unique(all_comets_no_debris$AnimalID)

for (animal_id in unique_animal_ids) {
  subset_data <- all_comets_no_debris[all_comets_no_debris$AnimalID == animal_id, ]
  
  fit <- densityMclust(subset_data$total_intensity, modelNames = "V", G = 3, verbose = FALSE, plot = FALSE)
  
  SD1 <- sqrt(fit[["parameters"]][["variance"]][["sigmasq"]][[1]]) +
    fit[["parameters"]][["mean"]][[1]]
  
  SD2 <- 2 * sqrt(fit[["parameters"]][["variance"]][["sigmasq"]][[1]]) +
    fit[["parameters"]][["mean"]][[1]]
  
  keep1 <- subset_data %>%
    dplyr::filter(total_intensity < SD1) %>%
    dim()
  keep1 <- keep1[1]
  
  keep2 <- subset_data %>%
    dplyr::filter(total_intensity < SD2) %>%
    dim()
  keep2 <- keep2[1]
  
  
  plot(
    fit,
    what = "density",
    data = subset_data$total_intensity,
    breaks = 50,
    col = "blue",
    lwd = 3,
    xlab = "DNA content (Tot_Int)"
  )
  abline(v = fit[["parameters"]][["mean"]][[1]], col = "red")
  abline(v = SD1, col = "red")
  abline(v = SD2, col = "red")
  text(fit[["parameters"]][["mean"]][[1]],
       0.000001,
       "1st mean",
       col = "red",
       adj = c(0, -1))
  text(SD1,
       0.000003,
       "1st mean + 1SD",
       col = "red",
       adj = c(0, -1))
  text(SD2,
       0.000005,
       "1st mean + 2SD",
       col = "red",
       adj = c(0, -1))
  if (length(fit[["parameters"]][["mean"]]) > 1) {
    abline(v = fit[["parameters"]][["mean"]][[2]], col = "darkgreen")
    text(fit[["parameters"]][["mean"]][[2]],
         0.000009,
         "2nd mean",
         col = "darkgreen",
         adj = c(0, -.1))
  }
  if (length(fit[["parameters"]][["mean"]]) > 2) {
    abline(v = fit[["parameters"]][["mean"]][[3]], col = "orange")
    text(fit[["parameters"]][["mean"]][[3]],
         0.000009,
         "3rd mean",
         col = "orange",
         adj = c(0, -.1))
  }
  title(main = paste0(animal_id, ": ", keep1, " and ", keep2, " comets when 1st mean + 1SD and 1st mean +2SD are used, respectively" ), cex.main = 0.8)
}


```

```{r}

## Export the 1C population data and >1C populations data (File names: “1C_modelling and More_than_1C_modelling”; median %TI, # comets) of each animal to a txt- file (play button #14).

# Renamed lists for clarity
comets_1C <- list()
comets_more_than_1C <- list()

for (animal_id in unique_animal_ids) {
  subset_data <- all_comets_no_debris[all_comets_no_debris$AnimalID == animal_id, ]
  
  fit <- tryCatch({
    densityMclust(subset_data$total_intensity, modelNames = "V", G = 3, verbose = FALSE, plot = FALSE)
  }, error = function(e) NULL)
  
  if (!is.null(fit)) {  
    mean_total_intensity <- fit$parameters$mean[1]
    SD1 <- sqrt(fit[["parameters"]][["variance"]][["sigmasq"]][[1]]) + mean_total_intensity
    SD2 <- 2 * sqrt(fit[["parameters"]][["variance"]][["sigmasq"]][[1]]) + mean_total_intensity
    
    comets_1C_1SD <- subset_data[subset_data$total_intensity < SD1, ]
    comets_1C_2SD <- subset_data[subset_data$total_intensity < SD2, ]
    

    comets_more_than_1C_1SD <- subset_data[subset_data$total_intensity >= SD1, ]
    comets_more_than_1C_2SD <- subset_data[subset_data$total_intensity >= SD2, ]
    

    comet_counts_1C_1SD <- nrow(comets_1C_1SD)
    median_tail_intensity_1C_1SD <- median(comets_1C_1SD$tail_intensity)
    comet_counts_1C_2SD <- nrow(comets_1C_2SD)
    median_tail_intensity_1C_2SD <- median(comets_1C_2SD$tail_intensity)
    
    
    comet_counts_more_than_1C_1SD <- nrow(comets_more_than_1C_1SD)
    median_tail_intensity_more_than_1C_1SD <- ifelse(comet_counts_more_than_1C_1SD > 0, median(comets_more_than_1C_1SD$tail_intensity), NA)
    
    
    comet_counts_more_than_1C_2SD <- nrow(comets_more_than_1C_2SD)
    median_tail_intensity_more_than_1C_2SD <- ifelse(comet_counts_more_than_1C_2SD > 0, median(comets_more_than_1C_2SD$tail_intensity), NA)
    
    
    animal_fit_data <- data.frame(
      AnimalID = animal_id,
      Mean_Total_Intensity_peak1 = mean_total_intensity,
      Comets_within_1SD = comet_counts_1C_1SD,
      Median_Tail_Intensity_within_1SD = median_tail_intensity_1C_1SD,
      Comets_within_2SD = comet_counts_1C_2SD,
      Median_Tail_Intensity_within_2SD = median_tail_intensity_1C_2SD
    )
    
    
    animal_filtered_out_data <- data.frame(
      AnimalID = animal_id,
      Comets_more_than_1SD = comet_counts_more_than_1C_1SD,
      Median_Tail_Intensity_more_than_1SD = median_tail_intensity_more_than_1C_1SD,
      Comets_more_than_2SD = comet_counts_more_than_1C_2SD,
      Median_Tail_Intensity_more_than_2SD = median_tail_intensity_more_than_1C_2SD
    )
    
    comets_1C[[length(comets_1C) + 1]] <- animal_fit_data
    comets_more_than_1C[[length(comets_more_than_1C) + 1]] <- animal_filtered_out_data
  }
}

comets_data_1C <- do.call(rbind, comets_1C)
comets_data_more_than_1C <- do.call(rbind, comets_more_than_1C)

Modelling_1C <- paste0("1C_modelling_", testitem, ".txt")
write.table(comets_data_1C, file = Modelling_1C, sep = "\t", row.names = FALSE, quote = FALSE)

Modelling_More_than_1C <- paste0("More_than_1C_modelling_", testitem, ".txt")
write.table(comets_data_more_than_1C, file = Modelling_More_than_1C, sep = "\t", row.names = FALSE, quote = FALSE)

```


